The Hajada Imputation Test

### 2020-05-30 article.Rmd
spelling

---
title: "The Hajada Imputation Test"
author: "Jakub Kosterna, Dawid Przybyliński & Hanna Zdulska"
date: "20/05/2020"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---

@@

There are many different ways of dealing with missing values. Depending on their quantity and properties, various methods turn out to be the best. The fact is, there is no one universal best method of imputation, and different problems require various solutions. Using one and the same solution for each dataset is certainly not the best option, but sometimes it is convenient or even necessary to focus on one particular method without much insight into its effect - for example, when time limits us, or when, due to the complexity of the algorithm or class of problems, we are forced to choose only one method.

@@

# Introduction + motivation

@@

1. *mode / median replace* - basic process which puts in a place of missing data median cells from their columns for numerical values and dominants for categorical ones.
2. *remove rows* - trivial solution of removing rows containing any missing data.
3. *'bogo' random replace* - simple algorithm replacing NA values with random numbers or factors from their columns.
4. *mice* ^1^ - advanced method creating multiple imputations for multivariate missing data, based on *Fully Conditional Specification* ^2^, where each incomplete variable is imputed by a separate model. Standard imputation with *maxit* parameter of value 0 and default 5 number of imputed datasets.
5. *VIM* ^3^ - standard k-nearest neighbors algorithm taken from the library.
6. *missForest* ^4^ - imputation based on random forest offered by the package.

@@

The fourteen data frames were taken from **OpenML100 collection** and were corrected specifically for this research. Both small and simple frames were chosen, as well as more complex and containing a lot of information, difficult to explain or present. They can be found under the following identifiers with the following names: 1590 - *adult*, 188 - *eucalyptus*, 23381 - *dresses-sales*, 29 - *credit-approval*, 38 - *sick*, 40536 - *SpeedDating*, 41278 - *okcupid-stem*, 56 - *vote*, 6332 - *cylinder-bands*, 1018 - *ipums_la_99-small*, 27 - *colic*, 4 - *labor*, 55 - *hepatitis* and 944 - *echoMonths*.

@@

The six imputations mentioned earlier were conducted on all fourteen of them, but only on six, imputations were successful in their entirety.

@@

In order to compare the quality of the imputed data and see how this supplementation of deficiencies works in practice, five binary classification algorithms have been selected. Their choice was made after careful analysis and extensive discussion, in order to find models that are both widely known and used, and also apply to different approaches and give reasonably distinguishable results using different techniques. The final choice fell on:

@@

Judging by the logarithmic scale, there is no surprise that removing rows and mode/median replenishment are definitely the fastest methods, with removing rows being several times faster. what would you expect, also "random values" imputation was quite fast, but slower than the two mentioned. Looking at more advanced ones, definitely *VIM* turned up to be the fastest - times usually being approximately 5-10 times shorter than *missForest* and *mice*. These last two are quite slow, with the former appearing to be slightly faster.

@@

Considering the imputation times and results, it can be said that the knn algorithm from the VIM package proved to be "universally" the best. Out of the advanced methods it falls out the fastest, and its results are also the best. However, it is more difficult to choose second and subsequent places - looking at the results, the simple methods are very quick, but they differ from the more advanced ones. Looking at the advanced ones, with the exception of VIM, MissForest stands out. However, his problem is time. So after VIM, it seems second best to use mean / median replenishment for a small amount of time and missForest for a large one. However, although mode/medain usually obtained worse ranks, there was no extensive difference in results between the methods - which is not visible in the charts. "removeRows" is characterized by large fluctuations, and therefore we advise against using it.

@@

The experiment can be considered successful when it comes to the datasets and tools made available to us - calculated measures, visualizations made, as well as conclusions led us, among others, to the intuition that probably *VIM* is really a resplendent package, supplementing the missing values with medians from columns is not is such an unwise idea as it may seem and *missForest* is also noteworthy - if we have time for running it and the dataset isn't too sizeable. Unfortunately, it is impossible to draw an objective conclusion based on such a small number of tested data - with as few as 14 sets, problems with imputations on some of them or too small dataframes not giving a satisfactory answer. Nevertheless, the code is certainly very valuable and for more datasets it could confirm one or another belief.

>>>>

---
title: "The Hajada Imputation Test"
author: "Jakub Kosterna, Dawid Przybyliński, Hanna Zdulska"
date: "May 2020"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---

@@

There are many different ways of dealing with missing values. Depending on their quantity and properties, various methods turn out to be the best. The fact is, there is no one universal best method of imputation, and different problems require various solutions. Using one and the same solution for each dataset is certainly not the best option, but sometimes it is convenient or even necessary to focus on one particular method without much insight into its effect - for example, when time limits us or when, due to the complexity of the algorithm or class of problems, we are forced to choose only one method.

@@

# Introduction and motivation

@@

1. *mode / median replace* - basic process which puts in a place of missing data median cells from their columns for numerical values and dominants for categorical ones.
2. *remove rows* - trivial solution of removing rows containing any missing data.
3. *'bogo' random replace* - simple algorithm replacing NA values with random numbers or factors from their features.
4. *mice* ^1^ - advanced method creating multiple imputations for multivariate missing data, based on *Fully Conditional Specification* ^2^, where each incomplete variable is imputed by a separate model. Standard imputation with *maxit* parameter of value 0 and default 5 number of imputed datasets.
5. *VIM* ^3^ - standard k-nearest neighbors algorithm taken from the library.
6. *missForest* ^4^ - imputation based on random forest offered by the package.

@@

The fourteen data frames were taken from **OpenML100 collection** and were corrected specifically for this research. Both small and simple frames were chosen as well as more complex and containing a lot of information, difficult to explain or present. They can be found under the following identifiers with the following names: 1590 - *adult*, 188 - *eucalyptus*, 23381 - *dresses-sales*, 29 - *credit-approval*, 38 - *sick*, 40536 - *SpeedDating*, 41278 - *okcupid-stem*, 56 - *vote*, 6332 - *cylinder-bands*, 1018 - *ipums_la_99-small*, 27 - *colic*, 4 - *labor*, 55 - *hepatitis* and 944 - *echoMonths*.

@@

The six imputations mentioned earlier were conducted on all fourteen of them, but only on six imputations were successful in their entirety.

@@

In order to compare the quality of the imputed data and see how this supplementation of deficiencies works in practice, five binary classification algorithms have been selected. Their choice was made after careful analysis and extensive discussion, in order to find models that are both widely known and used, but also apply to different approaches and give reasonably distinguishable results using different techniques. The final choice fell on:

@@

Judging by the logarithmic scale, there is no surprise that removing rows and mode/median replenishment are definitely the fastest methods, with removing rows being several times faster. What would you expect, also "random values" imputation was quite fast, but slower than the two mentioned. Looking at more advanced ones, definitely *VIM* turned up to be the fastest - times usually being approximately 5-10 times shorter than *missForest* and *mice*. These last two are quite slow, with the former appearing to be slightly faster.

@@

Considering the imputation times and results, it can be said that the knn algorithm from the VIM package proved to be "universally" the best. Out of the advanced methods it falls out the fastest, and its results are also the best. However, it is more difficult to choose second and subsequent places - looking at the results, the simple methods are very quick, but they differ from the more advanced ones. Looking at the advanced ones, with the exception of VIM, MissForest stands out. However, his problem is time. So after VIM, it seems second best to use mean / median replenishment for a small amount of time and missForest for a large one. However, although mode / medain usually obtained worse ranks, there was no extensive difference in results between the methods - which is not visible in the charts. "removeRows" is characterized by large fluctuations, and therefore we advise against using it.

@@

The experiment can be considered successful when it comes to the datasets and tools made available to us - calculated measures, visualizations made, as well as conclusions led us, among others, to the intuition that probably *VIM* is really a resplendent package, supplementing the missing values with medians from columns is not is such an unwise idea as it may seem and *missForest* is also noteworthy - if one have time for running it and the dataset isn't too sizeable. Unfortunately, it is impossible to draw an objective conclusion based on such a small number of tested data - with as few as 14 sets, problems with imputations on some of them or too small dataframes not giving a satisfactory answer. Nevertheless, the code is certainly very valuable and for more datasets it could confirm one or another belief.
